\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-02-12 move_schedules.tex
Updates from Overleaf

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract}
\abstract{
	The standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices is the reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications.
%	As our first result, we propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that takes $O(m^2)$ preprocessing time uses only $O(m)$ space. 
%	We also show that this maintenance procedure yields a factor-2 net reduction in column operations over the state of the art approach. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that requires only $O(m \log \log m)$ time and $O(m)$ space to construct. 
By reduction to a particular longest common subsequence problem, we show the storage needed to employ this strategy is actually sublinear in expectation. 
Exploiting this connection, we show experimentally the decrease in operations to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
}
%\end{abstract}

@@

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

@@

\subsection{Related Work}\label{sec:related_work} 
To the authors knowledge, work focused on ways of updating a  decomposition $R = DV$, for all homological dimensions, is limited: there is the vineyards algorithm~\cite{cohen2006vines} and the moves algorithm~\cite{busaryev2010tracking}, both of which are discussed extensively in section~\ref{sec:background}. At the time of writing, we were made aware of very recent work~\cite{luo2021accelerating} that iteratively repairs a permuted decomposition via a column swapping and reduce strategy, which they call ``warm starts.'' Though similar in spirit, their approach relies on the reduction algorithm as a subprocedure, which is quite different from the strategy we employ here.

@@

Each image contains a fixed number of pixels whose intensities vary with time. For each image, we build a fixed-size filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane whose inclusion maps are obtained by lower stars of pixel values.   
%Snapshots of the image data and its corresponding time-varying filtration are shown below Figure~\ref{fig:vineyards}. 
%\begin{figure*}[h!]
%\centering	
%\end{figure*}
%Note that the size of the filtration is fixed---no simplices are added or removed over time. 
Two events critically change the persistence diagrams: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus  splits into four components. 
From left to right, the Betti numbers of the five evenly spaced `snapshots' of the filtration shown above are: $(\beta_0,\beta_1)  = (1,0), \; (1,1),  \; (1,1), \; (1,1), \; (4,0)$.
Thus, in this example, only a few persistence diagrams are needed to capture the major changes to the topology.

@@

Note that without knowing where persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when the pairing function determining the persistence diagram is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations for dynamic persistence settings. 
By a reduction to an edit distance problem, we give a   lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition,  along with a proof of its optimality.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for constructing these operations---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
These operations parameterize sequences of cyclic permutations $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for heuristically solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

We now summarize the main results obtained in this effort (Theorem~\ref{prop:sparsity_result}): suppose one has a sequence of permutations, i.e. a schedule,  $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where each permutation $s_k = (i_k, j_k)$ is a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_\ast < j_\ast$ for all $s_\ast \in \mathcal{S}$, our first result extends~\cite{busaryev2010tracking} by showing that~\eqref{eq:rv_seq} can be computed  using $O(\kappa)$ column operations, where: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert 
\end{equation}
The quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $f$ and $f'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself and is thus output sensitive, which we take advantage of in Section~\ref{sec:schedule_cost}. 
%Surprisingly, due to a refined analysis of restore subprocedure from~\cite{busaryev2010tracking} we give in section~\ref{sec:continuous_setting}, we find that move operations are actually cheaper than vineyards, see Proposition~\ref{prop:factor2}.
%if the vineyards algorithm requires $T$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that obtains the same sequence using at most $M$ column operations, where: 
%\begin{equation}
%	M \leq \tfrac{1}{2}  T
%\end{equation}

@@

By reinterpreting a special set of cyclic permutations as edit operations on strings, we find that the \emph{smallest} such sequence that maps $f \mapsto f'$ has size  (Proposition~\ref{prop:schedule_size}) : 

@@

We also show that the information needed to construct any $\mathcal{S}$ with optimal size can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$ clearly is $O(m)$ for pathological inputs, we give evidence  that $d \sim m - \sqrt{m}$ in expectation for random filtrations is not too pessimistic for truly random filtrations (Corollary~\ref{cor:expectation}), and we give empirical results suggesting $d$ is tends to be much smaller for time-varying filtrations. 

@@

In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 

@@

\section{Background}\label{sec:background} 
%\subsection{Notation}\label{sec:notation} 
Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. 
Such a family is called a \emph{filtration},
which is deemed \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into an simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
Equivalently, we can also define a filtration as a pair $(K, f)$ where $K$ is a simplicial complex and $f : K \to I$ is a \emph{filter function} over a totally ordered index set $I$ satisfying $f(\tau) \leq f(\sigma)$ whenever $\tau \subseteq \sigma$, for any $\tau,\sigma \in K$. In this setting, $K_i = \{ \, \sigma \in K :  f(\sigma) \leq i \, \}$. Here, we consider two index sets: $[m] = \{ 1, \dots, m\}$ and $\mathbb{R}$. 
Without loss of generality, we exclusively consider simplexwise filtrations but for brevity-sake refer to them simply as filtrations.

Let $K$ be an abstract simplicial complex and $\mathbb{F}$ a field.
A $p$-chain is a
formal $\mathbb{F}$-linear combination  of $p$-simplices of $K$. The collection of $p$-chains under addition yields an 
$\mathbb{F}$-vector space  denoted   $C_p(K)$. 
The $p$-boundary $\partial_p(\sigma)$  of a $p$-simplex $\sigma\in K$ is the alternating sum of its oriented co-dimension 1 faces,
and the $p$-boundary of a $p$-chain is defined 
linearly in terms of its constitutive simplices. 
A $p$-chain with zero boundary is called a $p$-cycle, and together they form $Z_p(K) = \mathrm{Ker}\,\partial_p$. 
Similarly, the collection of $p$-boundaries forms  $B_p(K) = \mathrm{Im}\,\partial_{p+1}$.
Since $\partial_p \circ \partial_{p+1} = 0$ for all $p\geq 0$, 
then the quotient space $H_p(K) = Z_p(K) / B_{p}(K)$ is well-defined, and called the 
$p$-th homology of $K$ with coefficients in $\mathbb{F}$. 
If $ f: K \rightarrow  [m]$ is a filtration, then the inclusion maps  $K_i\subseteq K_{i+1}$   induce linear transformations 
%$f_p^{i,j}: H_p(K_i) \to H_p(K_j)$ 
at the level of homology:
\begin{equation}
	H_p(K_1) \to H_p(K_2) \to \dots \to H_p(K_m)
\end{equation}
%The $p$-th persistent homology groups are the images of these transformations: $H_{p}^{i,j} = \mathrm{Im}\,f_p^{i,j}$. 
%Note that if $i = j$, then $H_{p}^{i,j} = H_{p}(K_i) = H_{p}(K_i)$ is   just the ``standard'' homology. 
Simplices whose inclusion in the filtration creates a new homology class are   called \emph{creators}, and simplices that destroy homology classes are   called \emph{destroyers}. 
The filtration indices of 
these creators/destroyers are referred to as \emph{birth} and \emph{death} times, respectively. 
The collection of birth/death  pairs 
$(i,j)$ is denoted $\mathrm{dgm}_p(K ,f )$, 
and referred to as the $p$-th \emph{persistence diagram} of $(K,f)$.
If a homology class is born at time $i$ and dies entering time $j$, the difference $\lvert i - j \rvert$ is called the \emph{persistence} of that class.
In practice, filtrations often arise from triangulations parameterized by geometric scaling parameters, and the ``persistence'' of a homology class actually refers to its lifetime with respect to the scaling parameter. 
%For this reason, we'll refer to the difference $\lvert i - j \rvert$ of a homology class $[c]$ as its \emph{index persistence}.
%\\
%\\
%\noindent \textbf{Example 2.1:} Non-simplexwise filtrations often arise in geometrical contexts. For example, given a finite metric space $(X, d)$, the \emph{Vietoris-Rips} complex at scale $\epsilon \in \mathbb{R}$ is the abstract simplicial complex given by: 
%\begin{equation}\label{eq:rips}
%	\mathrm{Rips_\epsilon}(X) = \{ S \subseteq X : S \neq \emptyset \;\mbox{ and }\;\mathrm{diam}(S) \leq \epsilon \}
%\end{equation} 
%The corresponding filtration is called the \emph{Vietoris-Rips} (VR) filtration, indexed by $I = \mathbb{R}$. 
%A VR filtration indexed over $\mathbb{R}$ can be condensed to an simplexwise filtration indexed over the set of pairwise distances $\{\, d(x,x') \; \mid \; x,x' \in X \, \}$  by extending a total order of $X$ to its power set using any linear extension of the face poset, such as e.g. the lexicographical ordering.

Let $\mathbb{X}$ be a triangulable topological space and $K$ an abstract simplicial complex whose geometric realization is homeomorphic to $\mathbb{X}$. 
Let  $f: \mathbb{X} \to \mathbb{R}$ be continuous and  write $\mathbb{X}_a = f^{-1}(-\infty, a]$ to denote the sublevel sets of $\mathbb{X}$ defined by the value $a$. 
A \emph{homological critical value} of $f$ is any value $a \in \mathbb{R}$ such that the homology of the sublevel sets of $f$ changes at $a$, i.e.  if for some $p$ the inclusion-induced homomorphism  $H_p(\mathbb{X}_{a - \epsilon}) \to H_p(\mathbb{X}_{a+\epsilon})$ is not an isomorphism for any small enough $\epsilon >0$. If there are only finitely many of these homological critical values, then $f$ is said to be \emph{tame}. The concept of homological critical points and tameness will be revisited in section~\ref{sec:vineyards}.

@@

 Given a filtration $(K, f)$ with $m$ simplices,  the main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagram is encoded in $R$ and the generating cycles in the columns of $V$.
 To begin the reduction, one starts by assembling the elementary boundary chains $\partial(\sigma)$ as columns ordered according to $f$ into a $m \times m$  \emph{filtration boundary matrix} $D$. Setting $V = I$ and $R = D$, one proceeds by performing elementary left-to-right column operations on $V$ and $R$ until the following invariants are satisfied:
\textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}
 	\item[I1.] $R = D V$ where $D$ is the boundary matrix of the filtration $K$
 	\item[I2.] $V$ is full-rank upper-triangular
 	\item[l3.] $R$ is \emph{reduced}: if $\mathrm{col}_i(R) \neq 0$ and $\mathrm{col}_j(R) \neq 0$, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 
 	\end{enumerate} 
 \vspace*{1em}
 where $\mathrm{low}_R(i)$ denotes the largest row index of a non-zero entry in column $i$ of $R$. 
We call the decomposition satisfying these three invariants \emph{valid}. Note that though the matrices $R$ and $V$ are not unique, the collection of persistent pairings are~\cite{zomorodian2005computing}. The persistence diagram of the corresponding filtration $K$ may be read off from lowest entries in $R$, once $R$ is in reduced form.

 It is at times more succinct to restrict to specific sub-matrices $D$ based on the homology dimension $p$, and so we write $D_p$ represent the $d_{p-1} \times d_p$ matrix representing the linear operator $\partial_p$ (the same notation is extended to $R$ and $V$).
We illustrate the reduction algorithm with an example pair $(R_1, V_1)$ below. 

@@

and whose filtration order is given as $K = (u, v, w, a, b, c)$. 

@@

Using $\mathbb{Z}_2$ coefficients, the reduction proceeds to compute $H_1(K)$ as follows:

@@

Despite this high algorithmic complexity, the number of column operations has been observed to be super-linear in practice, due impart to the high sparsity and structure of $D$. 

>>>>

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract}
\abstract{
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications.
%	As our first result, we propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that takes $O(m^2)$ preprocessing time uses only $O(m)$ space. 
%	We also show that this maintenance procedure yields a factor-2 net reduction in column operations over the state of the art approach. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that requires only $O(m \log \log m)$ time and $O(m)$ space to construct. 
By reduction to a particular longest common subsequence problem, we show the storage needed to employ this strategy is actually sublinear in expectation. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations, is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
}
%\end{abstract}

@@

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagrams of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

@@

\subsection{Related Work}\label{sec:related_work} 
To the authors knowledge, work focused on ways of updating a  decomposition $R = DV$, for all homological dimensions, is limited: there is the vineyards algorithm~\cite{cohen2006vines} and the moves algorithm~\cite{busaryev2010tracking}, both of which are discussed extensively in section~\ref{sec:background}. At the time of writing, we were made aware of very recent work~\cite{luo2021accelerating} that iteratively repairs a permuted decomposition via a column swapping and reduce strategy, which they call ``warm starts.'' Though their motivation is similar to our own, their approach relies on the reduction algorithm as a subprocedure, which is quite different from the strategy we employ here.

@@

Each image in the series is comprised of pixels whose intensities vary with time, upon which we build a simplicial complex using the \emph{Freudenthal} triangulation of the plane. For each complex, we create a filtration of simplices whose order is determined by the lower stars of pixel values.   
%Snapshots of the image data and its corresponding time-varying filtration are shown below Figure~\ref{fig:vineyards}. 
%\begin{figure*}[h!]
%\centering	
%\end{figure*}
%Note that the size of the filtration is fixed---no simplices are added or removed over time. 
Two events critically change the persistence diagrams: the first occurs when the central connected component splits to form a cycle, and the second  when the annulus  splits into four components. 
From left to right, the Betti numbers of the five evenly spaced `snapshots' of the filtration shown above are: $(\beta_0,\beta_1)  = (1,0), \; (1,1),  \; (1,1), \; (1,1), \; (4,0)$.
Thus, in this example, only a few persistence diagrams are needed to capture the major changes to the topology.

@@

Note that without knowing where the persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

The existence of a time discretization that is more efficient to compute than continually updating the decomposition, indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when the pairing function determining the persistence diagram is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations for dynamic persistence settings. 
By a reduction to an edit distance problem, we give a   lower bound on the minimal number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition,  along with a proof of its optimality.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for constructing these operations---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
These operations parameterize sequences of permutations $\mathcal{S}= ( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for heuristically solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

We now summarize our main results  (Theorem~\ref{prop:sparsity_result}): suppose one has a  schedule  $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where  $s_k = (i_k, j_k)$ for $k=1,\ldots, d$,  denotes a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_k < j_k$ for all $s_k \in \mathcal{S}$, our first result extends~\cite{busaryev2010tracking} by showing that~\eqref{eq:rv_seq} can be computed  using $O(\kappa)$ column operations, where: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert 
\end{equation}
The quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $f$ and $f'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself and is thus output sensitive, which we leverage in Section~\ref{sec:schedule_cost}. 
%Surprisingly, due to a refined analysis of restore subprocedure from~\cite{busaryev2010tracking} we give in section~\ref{sec:continuous_setting}, we find that move operations are actually cheaper than vineyards, see Proposition~\ref{prop:factor2}.
%if the vineyards algorithm requires $T$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that obtains the same sequence using at most $M$ column operations, where: 
%\begin{equation}
%	M \leq \tfrac{1}{2}  T
%\end{equation}

@@

By reinterpreting a special set of cyclic permutations as edit operations on strings, we find that a  \emph{smallest} such sequence mapping $f $ to $  f'$, has size  (Proposition~\ref{prop:schedule_size}) : 

@@

We also show that the information needed to construct any $\mathcal{S}$ with optimal size can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$  can be $O(m)$ for pathological inputs, we provide evidence  that $d \sim m - \sqrt{m}$ in expectation for random filtrations   (Corollary~\ref{cor:expectation}), and we give empirical results suggesting $d$  can be much smaller for time-varying filtrations. 

@@

In Section \ref{sec:move_schedules} we introduce move schedules and provide efficient algorithms to construct them. 

@@

\section{Background}\label{sec:background} 
%\subsection{Notation}\label{sec:notation} 
Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. 
Such a family is called a \emph{filtration},
which is deemed \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into an simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
Equivalently, a filtration can be also defined as a pair $(K, f)$ where $K$ is a simplicial complex and $f : K \to I$ is a \emph{filter function}  satisfying $f(\tau) \leq f(\sigma)$ in $I$, whenever $\tau \subseteq \sigma$ in  $  K$. In this setting, $K_i = \{ \, \sigma \in K :  f(\sigma) \leq i \, \}$. Here, we consider two index sets: $[m] = \{ 1, \dots, m\}$ and $\mathbb{R}$. 
Without loss of generality, we exclusively consider simplexwise filtrations, but for brevity-sake refer to them simply as filtrations.

Let $K$ be an abstract simplicial complex and $\mathbb{F}$ a field.
A $p$-chain is a
formal $\mathbb{F}$-linear combination  of $p$-simplices of $K$. The collection of $p$-chains under addition yields an 
$\mathbb{F}$-vector space  denoted   $C_p(K)$. 
The $p$-boundary $\partial_p(\sigma)$  of a $p$-simplex $\sigma\in K$ is the alternating sum of its oriented co-dimension 1 faces,
and the $p$-boundary of a $p$-chain is defined 
linearly in terms of its constitutive simplices. 
A $p$-chain with zero boundary is called a $p$-cycle, and together they form $Z_p(K) = \mathrm{Ker}\,\partial_p$. 
Similarly, the collection of $p$-boundaries forms  $B_p(K) = \mathrm{Im}\,\partial_{p+1}$.
Since $\partial_p \circ \partial_{p+1} = 0$ for all $p\geq 0$, 
then the quotient space $H_p(K) = Z_p(K) / B_{p}(K)$ is well-defined, and called the 
$p$-th homology of $K$ with coefficients in $\mathbb{F}$. 
If $ f: K \rightarrow  [m]$ is a filtration, then the inclusion maps  $K_i\subseteq K_{i+1}$   induce linear transformations 
%$f_p^{i,j}: H_p(K_i) \to H_p(K_j)$ 
at the level of homology:
\begin{equation}
	H_p(K_1) \to H_p(K_2) \to \dots \to H_p(K_m)
\end{equation}
%The $p$-th persistent homology groups are the images of these transformations: $H_{p}^{i,j} = \mathrm{Im}\,f_p^{i,j}$. 
%Note that if $i = j$, then $H_{p}^{i,j} = H_{p}(K_i) = H_{p}(K_i)$ is   just the ``standard'' homology. 
Simplices whose inclusion in the filtration creates a new homology class are   called \emph{creators}, and simplices that destroy homology classes are   called \emph{destroyers}. 
The filtration indices of 
these creators/destroyers are referred to as \emph{birth} and \emph{death} times, respectively. 
The collection of birth/death  pairs 
$(i,j)$ is denoted $\mathrm{dgm}_p(K ,f )$, 
and referred to as the $p$-th \emph{persistence diagram} of $(K,f)$.
If a homology class is born at   $K_i$ and dies entering   $K_j$, the difference $\lvert i - j \rvert$ is called the \emph{persistence} of that class.
In practice, filtrations often arise from triangulations parameterized by geometric scaling parameters, and the ``persistence'' of a homology class actually refers to its lifetime with respect to the scaling parameter. 
%For this reason, we'll refer to the difference $\lvert i - j \rvert$ of a homology class $[c]$ as its \emph{index persistence}.
%\\
%\\
%\noindent \textbf{Example 2.1:} Non-simplexwise filtrations often arise in geometrical contexts. For example, given a finite metric space $(X, d)$, the \emph{Vietoris-Rips} complex at scale $\epsilon \in \mathbb{R}$ is the abstract simplicial complex given by: 
%\begin{equation}\label{eq:rips}
%	\mathrm{Rips_\epsilon}(X) = \{ S \subseteq X : S \neq \emptyset \;\mbox{ and }\;\mathrm{diam}(S) \leq \epsilon \}
%\end{equation} 
%The corresponding filtration is called the \emph{Vietoris-Rips} (VR) filtration, indexed by $I = \mathbb{R}$. 
%A VR filtration indexed over $\mathbb{R}$ can be condensed to an simplexwise filtration indexed over the set of pairwise distances $\{\, d(x,x') \; \mid \; x,x' \in X \, \}$  by extending a total order of $X$ to its power set using any linear extension of the face poset, such as e.g. the lexicographical ordering.

Let $\mathbb{X}$ be a triangulable topological space; that is, so that there exists   an abstract simplicial complex $K$ whose geometric realization is homeomorphic to $\mathbb{X}$. 
Let  $f: \mathbb{X} \to \mathbb{R}$ be continuous and  write $\mathbb{X}_a = f^{-1}(-\infty, a]$ to denote the sublevel sets of $\mathbb{X}$ defined by the value $a$. 
A \emph{homological critical value} of $f$ is any value $a \in \mathbb{R}$ such that the homology of the sublevel sets of $f$ changes at $a$, i.e.  if for some $p$ the inclusion-induced homomorphism  $H_p(\mathbb{X}_{a - \epsilon}) \to H_p(\mathbb{X}_{a+\epsilon})$ is not an isomorphism for any small enough $\epsilon >0$. If there are only finitely many of these homological critical values, then $f$ is said to be \emph{tame}. The concept of homological critical points and tameness will be revisited in section~\ref{sec:vineyards}.

@@

 Given a filtration $(K, f)$ with $m$ simplices,  the main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagrams are encoded in $R$ and the generating cycles in the columns of $V$.
 To begin the reduction, one first   assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to $f$ into a $m \times m$  \emph{filtration boundary matrix} $D$. Setting $V = I$ and $R = D$, one proceeds by performing elementary left-to-right column operations on $V$ and $R$ until the following invariants are satisfied:
\textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}
 	\item[I1.] $R = D V$ where $D$ is the boundary matrix of the filtration $(K,f)$
 	\item[I2.] $V$ is full-rank upper-triangular
 	\item[l3.] $R$ is \emph{reduced}: if $\mathrm{col}_i(R) \neq 0$ and $\mathrm{col}_j(R) \neq 0$, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 
 	\end{enumerate} 
 \vspace*{1em}
 where $\mathrm{low}_R(i)$ denotes the largest row index of a non-zero entry in column $i$ of $R$. 
We call the decomposition satisfying these three invariants \emph{valid}. Note that though the matrices $R$ and $V$ are not unique, the collection of persistent pairings are~\cite{zomorodian2005computing}. The persistence diagrams of the corresponding filtration  can be determined from the lowest entries in $R$, once it has been reduced.

 It is at times more succinct to restrict to specific sub-matrices of $D$ based on the homology dimension $p$, and so we write $D_p$ to represent the $d_{p-1} \times d_p$ matrix representing $\partial_p$ (the same notation is extended to $R$ and $V$).
We illustrate the reduction algorithm with an example   below. 

@@

and whose filtration order is given as $ (u, v, w, a, b, c)$. 

@@

Using $\mathbb{Z}_2$ coefficients, the reduction proceeds to compute $(R_1,V_1)$ as follows:

@@

Despite this high algorithmic complexity, the number of column operations has been observed to be super-linear in practice, due in part to the high sparsity and structure of $D$. 
